---
title: "IAPS MARCER Behavioral Results"
author: "Mengzhe Wei"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: journal
    toc: yes
    toc_float:
      collapsed: true
---

# MARCER IAPS behavior summary
Started 10/28/2024 by Mengzhe Wei. Preliminary behavioral data analysis for MARCER IAPS task. 

```{r setup, warning=FALSE, message=FALSE, echo = FALSE, eval = TRUE}
# set up
rm(list=ls());
options(warnPartialMatchDollar=TRUE);   # safety option
library("tidyverse");
library("DescTools");
library(lme4); 

# Read the data (update this with the correct file path and format)
in.path <- "C:/Users/Emotion/Documents/GIT-MW/IAPS/"
data <- read.csv(paste0(in.path,"iaps_ratings.csv"))
rating <- c("pos.rating","neg.rating")
image <- c("positive","neutral","negative")
trial.type <- c("LookNeg", "PosReap")
do.trim <- 0.1;
```

```{r basic stats, warning=FALSE, message=FALSE, echo = FALSE, eval = TRUE}
#Some sanity check stats
test <- data %>%
  group_by(sub.id) %>%
  summarise(age.group = first(age.group),
            MDD.status = first(MDD.status),
  ) %>%
  ungroup()
table(test$age.group)
table(test$MDD.status)

test <- data %>%
  group_by(sub.id) %>%
  summarise(
    age.group = first(age.group),
    MDD.status = first(MDD.status),
    cont.pos = mean(continuous.rating[valence == "positive"], na.rm = TRUE),
    cont.neu = mean(continuous.rating[valence == "neutral"], na.rm = TRUE),
    cont.neg = mean(continuous.rating[valence == "negative"], na.rm = TRUE),
    mean_pos_pos = mean(pos.rating[valence == "positive"], na.rm = TRUE),
    mean_neg_pos = mean(neg.rating[valence == "positive"], na.rm = TRUE),
    mean_pos_neu = mean(pos.rating[valence == "neutral"], na.rm = TRUE),
    mean_neg_neu = mean(neg.rating[valence == "neutral"], na.rm = TRUE),
    mean_pos_neg = mean(pos.rating[valence == "negative"], na.rm = TRUE),
    mean_neg_neg = mean(neg.rating[valence == "negative"], na.rm = TRUE),
  ) %>%
  ungroup()

data %>%
  group_by(MDD.status)%>%
  summarise(pos.pos = median(pos.rating[valence == "positive"], na.rm = TRUE),
            neg.pos = median(neg.rating[valence == "positive"], na.rm = TRUE),
            pos.neu= median(pos.rating[valence =="neutral"], na.rm = TRUE),
            neg.neu= median(neg.rating[valence == "neutral"], na.rm = TRUE),
            pos.neg= median(pos.rating[valence == "negative"], na.rm = TRUE),
            neg.neg= median(neg.rating[valence == "negative"], na.rm = TRUE),
            cont.pos = median(continuous.rating[valence == "positive"], na.rm = TRUE),
            cont.neu = median(continuous.rating[valence == "neutral"], na.rm = TRUE),
            cont.neg = median(continuous.rating[valence == "negative"], na.rm = TRUE))

data %>%
  group_by(age.group)%>%
  summarise(pos.pos = median(pos.rating[valence == "positive"], na.rm = TRUE),
            neg.pos = median(neg.rating[valence == "positive"], na.rm = TRUE),
            pos.neu= median(pos.rating[valence =="neutral"], na.rm = TRUE),
            neg.neu= median(neg.rating[valence == "neutral"], na.rm = TRUE),
            pos.neg= median(pos.rating[valence == "negative"], na.rm = TRUE),
            neg.neg= median(neg.rating[valence == "negative"], na.rm = TRUE),
            cont.pos = median(continuous.rating[valence == "positive"], na.rm = TRUE),
            cont.neu = median(continuous.rating[valence == "neutral"], na.rm = TRUE),
            cont.neg = median(continuous.rating[valence == "negative"], na.rm = TRUE))


test %>%
  group_by(MDD.status, age.group)%>%
  summarise(pos.pos = mean(mean_pos_pos),
            neg.pos = mean(mean_neg_pos),
            pos.neu= mean(mean_pos_neu),
            neg.neu= mean(mean_neg_neu),
            pos.neg= mean(mean_pos_neg),
            neg.neg= mean(mean_neg_neg),
            cont.pos = mean(cont.pos),
            cont.neu = mean(cont.neu),
            cont.neg = mean(cont.neg))

```

```{r function, warning=FALSE, message=FALSE, echo = FALSE, eval = TRUE}
# Function for graphing; pos & neg rating
make.plot <- function(do.group, do.variable) { # do.group <- "MDD status"; do.variable<-"image"; do.variable<-"trial";
  colors <- c("green", "blue") 
  shifts <- c(-0.15, 0.15)  # Calculate positions for boxplots
  if (do.group == "age") {
    group <- c("MA", "OA");}
  if (do.group == "MDD status") {
    group <- c("HC", "MDD");} 
  if (do.group == "all") {
    group <- "ALL"; }
  if (do.variable == "image"){
    variable <- image
  }
  if (do.variable == "trial"){
    variable <- trial.type
  }
  for (gid in 1:length(group)) { # gid<-1;
    plot(x=0, y=0, xlim=c(0.5, length(variable) +0.5), ylim=c(1-0.05,7+0.05), xaxt = 'n', ylab="", xlab="", main="", cex.axis=0.85, col='white');
    if (do.group == "age" | do.group == "MDD status") {
      if (do.group == "age") {
        ttl <- paste0(group[gid] , " Group - ", length(unique(data$sub.id[which(data$age.group == group[gid])])), " participants");
      }
      if (do.group == "MDD status") {
        ttl <- paste0(group[gid] , " Group - ", length(unique(data$sub.id[which(data$MDD.status == group[gid])])), " participants");
      }
      mtext(side=3, text=ttl, line=1.2, cex=0.7); }
    mtext(side=3, text=paste0("Positive and Negative Ratings by ", do.variable, " Type"), line=0.1, cex=0.85); 
    mtext(side=2, text="Rating", line=1.3, cex=0.7);
    axis(side = 1, at = seq(1, by = 1, length.out = length(variable)), labels = variable, cex.axis = 0.9);
    grid(col='darkgrey');
    for (rid in 1:length(rating)) {
      for (vid in 1:length(variable)) { # rid<-1; vid<-1
        if (do.group == "age") {
          tbl <- data[which(data$age.group == group[gid]),]}
        if (do.group == "MDD status"){
          tbl <- data[which(data$MDD.status == group[gid]),]}
        if (do.group == "all"){
          tbl <- data }
        if (do.variable == "image"){
          idx <- which(tbl$valence == variable[vid])
        }
        if (do.variable == "trial"){
          idx <- which(tbl$trial.type == variable[vid])
        } 
        # Extract the data for the current rating type
        vals <- tbl[idx, rating[rid]]
        vals <- vals[!is.na(vals)]
        boxplot(vals, at = vid + shifts[rid], col = colors[rid], border = 'black', add = TRUE,  notch=FALSE, xaxt = 'n', yaxt = 'n', bty = 'n', boxwex = 0.5, outpch = 1, outcol = adjustcolor(colors[rid], alpha = 0.5))
      }
    }
    if (do.group == "all"){
      legend(x = "bottomright",  # Adjust the first value for horizontal positioning
             legend = c("pos.rating", "neg.rating"), 
             fill = colors, border = colors, bty = "o", cex = 0.7, title = "Rating Type")
    }
  }
}
```

```{r function2, warning=FALSE, message=FALSE, echo=FALSE, eval=TRUE}
make.cont.plot <- function(do.group,do.variable) { # do.group <- "MDD status"; do.group <- "all"; do.variable<-"image"; do.variable<-"trial";
  if (do.group == "age") {
    group <- c("MA", "OA"); shifts <- c(-0.15, 0.15); colors <- c("#008080", "darkblue");
  }
  if (do.group == "MDD status") {
    group <- c("HC", "MDD"); shifts <- c(-0.15, 0.15); colors <- c("#D62728", "#17BECF");
  }
  if (do.group == "all"){
    group <- "ALL"; shifts <- 0;colors <- "pink"}
  if (do.variable == "image"){
    variable <- image
  }
  if (do.variable == "trial"){
    variable <- trial.type
  }
  plot(x=0, y=0, xlim=c(0.5, length(variable) +0.5), ylim=c(-6-0.05,6+0.05), xaxt = 'n', ylab="", xlab="", main="", cex.axis=0.85, col='white');
  ttl<-paste0(do.group);
  mtext(side=3, text=paste0("Continuous Rating by ", do.variable, " Type"), line=0.1, cex=0.85); 
  mtext(side=3, text=ttl, line=1.2, cex=0.7); 
  mtext(side=2, text="Continuous Rating", line=1.3, cex=0.7);
  axis(side = 1, at = seq(1, by = 1, length.out = length(variable)), labels = variable, cex.axis = 0.9);
  grid(col='darkgrey');
  abline(h = 0, col = "darkgrey", lwd = 1) 
  for (gid in 1:length(group)) {  # gid <- 1;
    for (vid in 1:length(variable)) { # vid<-1;
      if (do.group == "age") {
        tbl <- data[which(data$age.group == group[gid]),]}
      if (do.group == "MDD status") {
        tbl <- data[which(data$MDD.status == group[gid]),]}
      if (do.group == "all"){
        tbl <- data }  
      if (do.variable == "image"){
          idx <- which(tbl$valence == variable[vid])
        }
        if (do.variable == "trial"){
          idx <- which(tbl$trial.type == variable[vid])
        }
      vals <- tbl[idx, "continuous.rating"]
      # Plot the boxplot
      boxplot(vals, at = vid+ shifts[gid], col =colors[gid], border = 'black', add = TRUE,  notch=FALSE, xaxt = 'n', yaxt = 'n', bty = 'n', boxwex = 0.5, outpch = 1)
    }
  }
}
```

## Positive, negative, and continuous ratings by image type
```{r image type ratings, warning=FALSE, echo = FALSE, message=FALSE,fig.height=3.5, fig.width=7.75, fig.align='center'}
#Positive and Negative Ratings
par(mar=c(3, 2.25, 2, 0.75), mgp=c(1.1, 0.2, 0), tcl=-0.2);
make.plot("all","image")
t<-YuenTTest(data$pos.rating[data$valence=="positive"], data$pos.rating[data$valence=="negative"], paired = FALSE, trim = do.trim);
t.value <- round(t$statistic, digits = 2);
p.value <- round(t$p.value, digits = 2);
t<-YuenTTest(data$neg.rating[data$valence=="positive"], data$neg.rating[data$valence=="negative"], paired = FALSE, trim = do.trim);
t.value <- round(t$statistic, digits = 2);
p.value <- round(t$p.value, digits = 2);

#Now divide by age group
layout(matrix(1:3, nrow = 1), widths = c(1, 1, 0.28))  # Add a third column for the legend
par(mar=c(3, 2.25, 2, 0.75), mgp=c(1.1, 0.2, 0), tcl=-0.2)
colors <- c("green", "blue") 
# Generate the two plots
make.plot("age","image") 
# Add a legend in the extra space on the right
par(xpd = TRUE)  # Allow plotting outside plot region
plot.new()
legend("center", legend = c("pos.rating", "neg.rating"), 
       fill = colors, border = colors, bty = "o", cex = 0.9, title = "Rating Type")

#Now divide by MDD group
layout(matrix(1:3, nrow = 1), widths = c(1, 1, 0.28))  # Add a third column for the legend
par(mar=c(3, 2.25, 2, 0.75), mgp=c(1.1, 0.2, 0), tcl=-0.2)
# Generate the two plots
make.plot("MDD status","image")
# Add a legend in the extra space on the right
par(xpd = TRUE)  # Allow plotting outside plot region
plot.new()
legend("center", legend = c("pos.rating", "neg.rating"), 
       fill = colors, border = colors, bty = "o", cex = 0.9, title = "Rating Type")

#Continuous rating
par(mar=c(3, 2.25, 2, 0.75), mgp=c(1.1, 0.2, 0), tcl=-0.2);
make.cont.plot("all","image")
t<-YuenTTest(data$continuous.rating[data$valence=="positive"], data$continuous.rating[data$valence=="negative"], paired = FALSE, trim = do.trim);
t.value <- round(t$statistic, digits = 2);
p.value <- round(t$p.value, digits = 2);

#gragh both age and mdd groups
layout(matrix(1:3, nrow = 1), widths = c(1, 1, 0.28))  # Add a third column for the legend
par(mar=c(3, 2.25, 2, 0.75), mgp=c(1.1, 0.2, 0), tcl=-0.2)
make.cont.plot("age","image")
make.cont.plot("MDD status","image")
par(xpd = TRUE)  # Allow plotting outside plot region
plot.new()
colors <- c("#008080", "darkblue", "#D62728", "#17BECF");
legend("center", legend = c("MA", "OA","MDD","HC"), 
       fill = colors, border = colors, bty = "o", cex = 0.9, title = "Participants")
```

## Personal Rating and Normative Rating
```{r normative, warning=FALSE, message=FALSE,echo = FALSE,fig.height=3.5, fig.width=7.75, fig.align='center'}
model <- lmer(continuous.rating ~ valence.overall + (1 | sub.id) + (1 | image),
              data = data)
summary(model)
model <- lmer(continuous.rating ~ valence.overall*MDD.status + (1 | sub.id) + (1 | image),
              data = data)
summary(model)
model <- lmer(continuous.rating ~ valence.overall*age.group + (1 | sub.id) + (1 | image),
              data = data)
summary(model)
ggplot(data, aes(x = valence.overall, y = continuous.rating, color = MDD.status)) +
  geom_point(alpha = 0.6) +  # Add points with some transparency
  geom_smooth(method = "lm", se = TRUE) +  # Add linear trend line with confidence interval
  labs(title = "Relationship Between Normative Valence and Continuous Rating",
       x = "Normative Valence (valence.overall)",
       y = "Continuous Rating") +
  theme_minimal()
ggplot(data, aes(x = valence.overall, y = continuous.rating, color = age.group)) +
  geom_point(alpha = 0.6) +  # Add points with some transparency
  geom_smooth(method = "lm", se = TRUE) +  # Add linear trend line with confidence interval
  labs(title = "Relationship Between Normative Valence and Continuous Rating",
       x = "Normative Valence (valence.overall)",
       y = "Continuous Rating") +
  theme_minimal()
```

## Positive, negative, and continuous ratings by trial types
```{r trial type ratings, warning=FALSE,echo = FALSE, message=FALSE,fig.height=3.5, fig.width=7.75, fig.align='center'}
#Positve and Negative Ratings
par(mar=c(3, 2.25, 2, 0.75), mgp=c(1.1, 0.2, 0), tcl=-0.2);
make.plot("all","trial")

#Now divide by age group
layout(matrix(1:3, nrow = 1), widths = c(1, 1, 0.28))  # Add a third column for the legend
par(mar=c(3, 2.25, 2, 0.75), mgp=c(1.1, 0.2, 0), tcl=-0.2)
make.plot("age","trial")
par(xpd = TRUE)  # Allow plotting outside plot region
plot.new()
legend("center", legend = c("pos.rating", "neg.rating"), 
       fill = colors, border = colors, bty = "o", cex = 0.9, title = "Rating Type")

#Now divide by MDD group
layout(matrix(1:3, nrow = 1), widths = c(1, 1, 0.28))  # Add a third column for the legend
par(mar=c(3, 2.25, 2, 0.75), mgp=c(1.1, 0.2, 0), tcl=-0.2)
make.plot("MDD status","trial")
par(xpd = TRUE)  # Allow plotting outside plot region
plot.new()
legend("center", legend = c("pos.rating", "neg.rating"), 
       fill = colors, border = colors, bty = "o", cex = 0.9, title = "Rating Type")

#Continuous ratings
par(mar=c(3, 2.25, 2, 0.75), mgp=c(1.1, 0.2, 0), tcl=-0.2);
make.cont.plot("all","trial")
#gragh both age and mdd groups
colors <- c("#008080", "darkblue", "#D62728", "#17BECF");
layout(matrix(1:3, nrow = 1), widths = c(1, 1, 0.28))  # Add a third column for the legend
par(mar=c(3, 2.25, 2, 0.75), mgp=c(1.1, 0.2, 0), tcl=-0.2)
make.cont.plot("age","trial")
make.cont.plot("MDD status","trial")
par(xpd = TRUE)  # Allow plotting outside plot region
plot.new()
legend("center", legend = c("MA", "OA","MDD","HC"), 
       fill = colors, border = colors, bty = "o", cex = 0.9, title = "Participants")
```

## Correlation between positive and negative rating (subject-level)
```{r pos&neg rating subject-level, warning=FALSE,echo = TRUE, message=FALSE}
#Only counting confirmed ratings. In previous code unconfirmed ratings were used too.
data_test <- data %>%
  group_by(sub.id) %>%
  summarize(
    mean.pos = mean(pos.rating[pos.confirmed == TRUE], na.rm = TRUE),
    mean.neg = mean(neg.rating[neg.confirmed == TRUE], na.rm = TRUE),
    sub.id = first(sub.id),
    age.group = first(age.group),
    MDD.status = first(MDD.status)
  ) %>%
  ungroup()

#all subject correlation
cor(data_test$mean.pos, data_test$mean.neg, use = "complete.obs")
# -0.032, very correlation

#subgroup
cor(data_test$mean.pos[data_test$MDD.status == "MDD"], data_test$mean.neg[data_test$MDD.status == "MDD"], use = "complete.obs")
#-0.028, very correlation
cor(data_test$mean.pos[data_test$MDD.status == "HC"], data_test$mean.neg[data_test$MDD.status == "HC"], use = "complete.obs")
#-0.037, very weak correlation
cor(data_test$mean.pos[data_test$age.group == "OA"], data_test$mean.neg[data_test$age.group == "OA"], use = "complete.obs")
#-0.185, weak correlation; quite high compared to the other groups. Shown on the graph below, it looks like a trend particular to OAHC group.
cor(data_test$mean.pos[data_test$age.group == "MA"], data_test$mean.neg[data_test$age.group == "MA"], use = "complete.obs")
#0.01, very weak correlation

colors <- c(MAHC = "#008080", OAHC = "darkblue", MAMDD = "#D62728", OAMDD = "#17BECF")
group_colors <- ifelse(data_test$age.group == "MA"& data_test$MDD.status == "HC", colors["MAHC"],
                 ifelse(data_test$age.group == "OA"& data_test$MDD.status == "HC", colors["OAHC"],
                 ifelse(data_test$age.group == "MA"& data_test$MDD.status == "MDD", colors["MAMDD"], colors["OAMDD"])))
plot(
  data_test$mean.pos, 
  data_test$mean.neg, 
  xlab = "Mean Positive Ratings", 
  ylab = "Mean Negative Ratings", 
  main = "Positive vs. Negative Ratings by Group",
  pch = 16, 
  col = group_colors
)
groups <- expand.grid(MDD.status = unique(data_test$MDD.status), age.group = unique(data_test$age.group))
for (i in 1:nrow(groups)) {
  group <- groups[i, ]
  subgroup_data <- data_test[data_test$MDD.status == group$MDD.status & data_test$age.group == group$age.group, ]
  if (nrow(subgroup_data) > 1) { # Ensure there are enough data points for regression
    abline(
      lm(mean.neg ~ mean.pos, data = subgroup_data),
      col = colors[paste0(group$age.group, group$MDD.status)],
      lwd = 2
    )
  }
}
legend(
  "topright", 
  legend = c("MAHC", "OAHC", "MAMDD", "OAMDD"), 
  col = colors, 
  lty = 1, 
  lwd = 2, 
  pch = 16
)
```
## Relationship between positive and negative rating (trial-wise)
```{r pos&neg rating trial-level, warning=FALSE, message=FALSE,echo = TRUE,fig.height=3.5, fig.width=7.75, fig.align='center'}
data_confirm <- data[data$pos.confirmed == TRUE & data$neg.confirmed == TRUE, ]
model_age <- lmer(neg.rating ~ pos.rating*age.group + (1 | sub.id),
              data = data_confirm)
summary(model_age)
# interesting age related difference here
model_MDD <- lmer(neg.rating ~ pos.rating*MDD.status + (1 | sub.id),
              data = data_confirm)
summary(model_MDD)
```

## Relationship between continuous rating and Reappraisal (trial-wise)
```{r reap effect, warning=FALSE, message=FALSE,echo = TRUE,fig.height=3.5, fig.width=7.75, fig.align='center'}
data_reap <- data_confirm[data_confirm$trial.type %in% c("LookNeg", "PosReap"), ]
model <- lmer(continuous.rating ~ trial.type + (1 | sub.id),
              data = data_reap)
summary(model)
#We are seeing a reappraisal effect
model_age <- lmer(continuous.rating ~ trial.type*age.group + (1 | sub.id),
              data = data_reap)
summary(model_age)
# Quite interesting of a direction. It seems like the reappraisal effect is weakened for OA
model_MDD <- lmer(continuous.rating ~ trial.type*MDD.status + (1 | sub.id),
              data = data_reap)
summary(model_MDD)
# The reappraisal effect is strengthened in MDD
```